Disentangling linkages between satellite derived forest structure and productivity across a range of forest types and ecosystems.

### 2023-03-20 scripts/manuscript/manuscript.qmd
first revisions to ncc 2023-03-17

---
title: Disentangling linkages between satellite derived forest structure and productivity across a range of forest types and ecosystems.
author:
  - name: Evan R. Muise
    email: evan.muise@student.ubc.ca
    affiliations: 
        - id: ubc
          name: University of British Columbia
          department: Forest Resources Management
          address: 2424 Main Mall
          city: Vancouver, BC, Canada
          postal-code: V6T 1Z4
    attributes:
        corresponding: true
  - name: Nicholas C. Coops
    email: nicholas.coops@ubc.ca
    affiliations: 
        - id: ubc
          name: University of British Columbia
          department: Forest Resources Management
          address: 2424 Main Mall
          city: Vancouver, BC, Canada
          postal-code: V6T 1Z4
  - name: Txomin Hermosilla
    email: txomin.hermosillagomez@NRCan-RNCan.gc.ca
    affiliations: 
        - id: cfs
          name: Natural Resources Canada
          department: Canada Forest Service (Pacific Forestry Centre)
          address: 506 Burnside Rd W
          city: Victoria, BC, Canada
          postal-code: V8Z 1M5
  - name: A. Cole Burton
    email: cole.burton@ubc.ca
    affiliations: 
        - id: ubc
          name: University of British Columbia
          department: Forest Resources Management
          address: 2424 Main Mall
          city: Vancouver, BC, Canada
          postal-code: V6T 1Z4
  - name: Margaret E. Andrew
    email: M.Andrew@murdoch.edu.au
    affiliations: 
        - id: murd
          name: Murdoch University
          department: Environmental and Conservation Sciences and Harry Butler Institute
          city: Murdoch, WA, Australia
          postal-code: 6150
  - name: Stephen S. Ban
    email: Stephen.Ban@gov.bc.ca
    affiliations: 
        - id: bcp
          name: Ministry of Environment and Climate Change Strategy
          department: BC Parks
          address: 525 Superior Street
          city: Victoria, BC, Canada
          postal-code: V8V 1T7
abstract: |
  insert abstract here
keywords: 
  - structural equation modelling
  - remote sensing
  - landsat
  - forest structure
  - forest productivity
date: last-modified
bibliography: bibliography.bib
csl: remote-sensing-of-environment.csl
format:
  elsevier-pdf:
    keep-tex: true
    keep-md: true
    journal:
      name: Remote Sensing of Environment
      formatting: review
      model: 3p
      cite-style: authoryear
      layout: twocolumn
  docx: 
    reference-doc: my-styles.docx
    keep-md: true
  pdf: 
    pdf-engine: xelatex
editor: 
  markdown: 
    wrap: 72
---

@@

There are six EBV classes, each corresponding to a different facet of
biodiversity, including, genetic composition, species populations,
species traits, community composition, ecosystem structure, and
ecosystem function [@pereira2013]. @fernández2020 divides the six
classes into two approaches, with one focusing on species biodiversity,
and the other focusing on ecosystem diversity. Remote sensing has proven
to be capable of measure five of the six classes, with the exception
being genetic composition, which requires in-situ observation and
samples [@skidmore2021]. Notably, while remote sensing can provide
information on the remaining two species EBV classes, this information
is typically acquired with an *ad-hoc* approach. This collects high
resolution, small spatial extent information, rather than the global or
regional scales required for biodiversity assessment. Community
composition falls into a similar dilemma, requiring species population
information which necessitates high resolution spatial data.

@@

Forest structural diversity has been linked to biodiversity at various
scales [@guo2017; @bergen2009; @gao2014]. Increased structural
complexity is hypothesized to create additional niches, leading to
increased species diversity [@bergen2009]. The relationship between
forest structure and biodiversity is commonly assessed using avian
species diversity metrics [@macarthur1961; @goetz2007], however, other
clades (and habitats), have been used [@davies2014; @nelson2005]. Many
metrics derived from lidar remote sensing have been used as local
indicators of biodiversity, including canopy cover, canopy height,
vertical profiles, and aboveground biomass, while other 2nd order
derived metrics such as canopy texture, height class distribution,
edges, and patch metrics have been used to examine habitat and
biodiversity at landscape scales [@bergen2009].

@@

Linkages between forest structure and productivity (namely, vegetation
indices) have been examined for nearly 20 years [@huete2002,
@knyazikhin1998; @myneni1994]. While there is significant theoretical
and empirical evidence for their relationship at single time points
(within a single image) [@myneni1994], various relationship directions
and shapes have been found between forest structure and productivity
metrics [@ali2019]. These relationships, their shapes, and their
strengths have been attributed to multiple possible hypotheses and can
vary based on environmental conditions [@ali2019]. The relationship
between forest structural diversity metrics and annual productivity
summaries has yet to be fully examined, including the DHIs.

Some vegetation structure metrics are simpler, and more accurate, to
calculate than others [@coops2021]. These basic metrics, such as canopy
height and canopy cover,can then be used to estimate additional
structure metrics, such as basal area or total biomass. Canopy height
and cover are commonly used as an indicator of vertical and horizontal
variation, respectively. Recently, more attention has been paid to
internal structural complexity metrics, which can be more difficult and
time consuming to generate [@coops2021; @ma2022] .

In this study, we seek to untangle the relationship between two EBVs:
forest structure diversity metrics and yearly summaries of forest
productivity. To accomplish this, we assess this relationship using path
analysis to assess the direct and indirect (as mediated by more complex
forest structural diversity metrics) effects of commonly collected
forest structural metrics (canopy height and canopy cover) on yearly
productivity summaries. Doing so will in turn assess their utility as
complementary EBVs. We ran this analysis separately across four forested
land covers and the forested ecosystems of British Columbia, Canada.

@@

British Columbia is the westernmost province of Canada, and is home to a
variety of terrestrial ecosystems [@pojar1987]. Approximately 64% of the
province is forested [@bcministryofforests2003]. There is a large amount
of ecosystem variation in the province, with large climate and
topographic gradients. The Biogeoclimatic Ecosystem Classification (BEC)
system identifies 16 zones based on the dominant tree species and the
ecosystems general climate. These zones can be further split into
subzones, variants, and phases based on microclimate, precipitation, and
topography.

@@

@@

@@

@@

>>>>

---
title: Disentangling linkages between satellite derived forest structure and productivity across a range of forest types and ecosystems.
author:
  - name: Evan R. Muise
    email: evan.muise@student.ubc.ca
    affiliations: 
        - id: ubc
          name: University of British Columbia
          department: Forest Resources Management
          address: 2424 Main Mall
          city: Vancouver, BC, Canada
          postal-code: V6T 1Z4
    attributes:
        corresponding: true
  - name: Margaret E. Andrew
    email: M.Andrew@murdoch.edu.au
    affiliations: 
        - id: murd
          name: Murdoch University
          department: Environmental and Conservation Sciences and Harry Butler Institute
          city: Murdoch, WA, Australia
          postal-code: 6150
  - name: Nicholas C. Coops
    email: nicholas.coops@ubc.ca
    affiliations: 
        - id: ubc
          name: University of British Columbia
          department: Forest Resources Management
          address: 2424 Main Mall
          city: Vancouver, BC, Canada
          postal-code: V6T 1Z4
  - name: Txomin Hermosilla
    email: txomin.hermosillagomez@NRCan-RNCan.gc.ca
    affiliations: 
        - id: cfs
          name: Natural Resources Canada
          department: Canada Forest Service (Pacific Forestry Centre)
          address: 506 Burnside Rd W
          city: Victoria, BC, Canada
          postal-code: V8Z 1M5
  - name: A. Cole Burton
    email: cole.burton@ubc.ca
    affiliations: 
        - id: ubc
          name: University of British Columbia
          department: Forest Resources Management
          address: 2424 Main Mall
          city: Vancouver, BC, Canada
          postal-code: V6T 1Z4
  - name: Stephen S. Ban
    email: Stephen.Ban@gov.bc.ca
    affiliations: 
        - id: bcp
          name: Ministry of Environment and Climate Change Strategy
          department: BC Parks
          address: 525 Superior Street
          city: Victoria, BC, Canada
          postal-code: V8V 1T7
abstract: |
  [insert abstract here - nicholas ignore the formatting on the authors for now, i can output to pdf and have it formatted nicely for submission]
keywords: 
  - structural equation modelling
  - remote sensing
  - landsat
  - forest structure
  - forest productivity
  - dynamic habitat indices
date: last-modified
bibliography: bibliography.bib
csl: remote-sensing-of-environment.csl
format:
  elsevier-pdf:
    keep-tex: true
    keep-md: true
    journal:
      name: Remote Sensing of Environment
      formatting: review
      model: 3p
      cite-style: authoryear
      layout: twocolumn
  docx: 
    reference-doc: my-styles.docx
    keep-md: true
  pdf: 
    pdf-engine: xelatex
editor: 
  markdown: 
    wrap: 72
execute: 
  echo: false
fig-width: 300
---

@@

There are six EBV classes, each corresponding to a different facet of
biodiversity, including, genetic composition, species populations,
species traits, community composition, ecosystem structure, and
ecosystem function [@pereira2013]. @fernández2020 divides the six
classes into two approaches, with one focusing on species biodiversity,
and the other focusing on ecosystem diversity. Remote sensing has proven
to be capable of measure five of the six classes, with the exception
being genetic composition, which requires in-situ observation and
samples [@skidmore2021]. Notably, while remote sensing can provide
information on the remaining two species EBV classes, this information
is typically acquired with an *ad-hoc* approach requiring high spatial
and spectral resolution data. This data can often only be collected at
local extents, rather than the global or regional scales required for
biodiversity trend assessment [@valdez2023]. Community composition falls
into a similar dilemma, requiring species population information which
necessitates high resolution spatial data.

@@

Forest structural diversity has been linked to biodiversity at various
scales [@guo2017; @bergen2009; @gao2014]. Increased structural
complexity is hypothesized to create additional niches, leading to
increased species diversity [@bergen2009]. The relationship between
forest structure and biodiversity is commonly assessed using avian
species diversity metrics [@macarthur1961; @goetz2007], however, other
clades (and habitats), have also been used [@davies2014; @nelson2005].
Many metrics derived from lidar remote sensing have been used as local
indicators of biodiversity, including canopy cover, canopy height,
vertical profiles, and aboveground biomass, while other 2nd order
derived metrics such as canopy texture, height class distribution,
edges, and patch metrics have been used to examine habitat and
biodiversity at landscape scales [@bergen2009].

@@

Some vegetation structure metrics are simpler, and more accurate, to
calculate than others [@coops2021]. These basic metrics, such as canopy
height and canopy cover, can then be used to estimate additional
structure metrics, such as basal area or total biomass. Canopy height
and cover are commonly used as indicators of vertical and horizontal
variation, respectively. Recently, more attention has been paid to
internal structural complexity metrics, which can be more difficult and
time consuming to generate [@coops2021; @ma2022], but have been shown to
have stronger linkages with biodiversity [@guo2017], and productivity
[@ali2019].

Linkages between forest structure and productivity (namely, vegetation
indices) have been examined for nearly 20 years [@huete2002,
@knyazikhin1998; @myneni1994]. While there is significant theoretical
and empirical evidence for their relationship at single time points
(within a single image) [@myneni1994], various relationship directions
and shapes have been found between forest structure and productivity
metrics [@ali2019]. These relationships, their shapes, and their
strengths have been attributed to multiple possible hypotheses and have
been shown to vary based on environmental conditions [@ali2019]. The
relationship between forest structural diversity metrics and annual
productivity summaries has yet to be fully examined, including the DHIs.

Understanding the complementarity between potential EBVs is an integral
component of their creation [@skidmore2021]. Biodiversity has been shown
to be linked to both forest structure [@guo2017; @gao2014], and
productivity [@radeloff2019] at a range of scales. Research has
highlighted that the relationship between productivity and biodiversity
is reciprocal [@worm2003], making productivity a suitable biodiversity
indicator. In forested ecosystems, stand structural attributes have also
been shown to influence productivity, with a range of responses
dependent on environmental conditions [@ali2019].

Structural equation modelling and path analyses have been commonly used
in ecology to assess the causal effects behind various hypotheses
[@fan2016; @grace2010], including the relationship between forest
structure, functioning, and biodiversity [@ali2019] . In this paper we
seek to untangle the relationship between two EBVs: forest structure
diversity metrics and yearly summaries of forest productivity. To
accomplish this, we assess this relationship using path analysis to
assess the direct and indirect (as mediated by more complex forest
structural diversity metrics) effects of commonly collected forest
structural metrics (canopy height and canopy cover) on yearly
productivity summaries. Further, we use exploratory structural equation
modelling (ESEM) to identify the latent variables driving the
productivity metrics. We run this analysis both globally, and stratified
by the forested ecosystems of British Columbia, Canada, and compare
models using AIC scores and global fit measures across the ecosystems.
The results from this study will assess the linkages and complementarity
between two biodiversity indicators.

@@

British Columbia is the westernmost province of Canada, and is home to a
variety of terrestrial ecosystems [@pojar1987]. Approximately 64% of the
province is forested [@bcministryofforests2003]. There is a large amount
of ecosystem variation in the province, with large climate and
topographic gradients. The Biogeoclimatic Ecosystem Classification (BEC)
system identifies 16 zones based on the dominant tree species and the
ecosystems general climate. These zones can be further split into
subzones, variants, and phases based on microclimate, precipitation, and
topography [@pojar1987].

{insert figure showing ecosystems of interest} need to remake figure
showing ecosystems, highlighting that bunchgrass isn't included

{table showing elevations and climate ranges for each included bec zone}

@@

### NTEMS

### Dynamic Habitat Indices

The Dynamic Habitat Indices are a set of satellite remote sensing
derived productivity variables that summarize the cumulative amount of
available energy, the minimum available energy, and the variation in
available energy throughout a given year [@radeloff2019]. The DHIs have
previously been produced at a global extent using MODIS imagery
[@radeloff2019], however, recent studies have examined how these indices
can be constructed at a finer spatial resolution using multi-annual
Landsat imagery.

@@

We conducted model based sampling across the fifteen forest dominated
ecosystems found within British Columbia \[BC figure reference, I have
not made this figure yet\]. Samples were randomly selected within BEC
zones of interest alongside multiple criteria. Each sampled pixel had to
have a forested land cover class (coniferous, deciduous, mixed-wood, or
wetland-treed), and be surrounded by the same land cover class.
Additionally, each pixel had to have a coefficient of variation less
than 0.5 in the Lorey's height and canopy cover forest structural
metrics. These focal analyses were preprocessed in R and Python, and
randomly sampled across a masked suitability raster for each BEC zone
using the *sgsR* R package version 1.3.4 [@goodbody2023]. A maximum of
3000 samples were taken from each BEC zone.

\[note to ncc; I've stopped doing this sampling step as Meg said she
didn't feel it was necessary\]

The imputation method used to generate the forest structure layers leads
to many duplicated values as there are a limited number of samples to
predict from [@matasci2018]. A discussion of imputation approaches to
large area lidar attribute modelling can be found in @coops2021. To
account for duplicated input values in our modelling we averaged the
endogenous variables within each ecosystem for each uniquely imputed set
of forest attributes, forest type, and ecosystem. Due to variations in
ecosystem size, differing numbers of samples were generated for each
ecosystem. To account for this during model comparison, each ecosystem
and land cover sample set was downsampled to the minimum number of
samples for a single BEC zone (in this case, 170). For the forest type
models, we attempted to select equal numbers of observations from each
BEC zone. However, some forest types are rarely found in certain zones.
To account for this, if an equal number of samples from each BEC zone
were not available, the entire sample from that zone was taken. The
amount of samples taken from other zones was then increased in order to
collect a total number equal to the minimum number of samples in a
single BEC zones (170). To meet the normality assumptions of path
analysis, all variables were natural-log transformed and standardized,
as per @grace2016. Variables containing zeros were natural-log plus one
transformed.

@@

### Path Analysis

To determine the relationships between simple lidar derived forest
structural attributes, complex/derived forest structural attributes, and
forest productivity, we used path analysis to analyze two causal models
(@fig-pathdag). To determine primary drivers of the three DHIs, we will
summarize the predictors across ecosystems by counting the strongest
predictor in each BEC zone. This will determine if the primary driver
for each facet of the DHI is simple or complex, and allow us to assess
ecosystem differences.

```{r}
#| label: fig-pathdag
#| fig-cap: Proposed path diagrams
#| warning: false
#| echo: false

knitr::include_graphics(here::here("outputs", "path_dags.png"))
```

### Exploratory Structural Equation Modelling

Secondly, we will use exploratory structural equation modelling
[@marsh2020; @asparouhov2009] to identify latent forest structural
variables within the data, and use these latent variables to predict the
dynamic habitat indices. Exploratory structural equation modelling is a
combination of exploratory factor analysis (EFA) and structural equation
modelling, which relaxes the strict requirement of zero cross-loadings
in confirmatory factor analysis and allows for less strict measurement
models to be used. ESEM is used when it is known that there is a latent
structure to the data, but the specific indicator variables have not yet
been determined.

One advantage of ESEM is that it can create varying numbers of latent
variables. For our analysis, we first ran exploratory factor analysis on
all forest structural attributes both globally and by BEC zone with
between 1-4 potential latent variables. We chose the most parsimonious
EFA model with the lowest AIC scores for each ecosystem, leading to
varying numbers of latent variables. We then determined the anchoring
indicator for each latent variable. The anchors were calculated by
choosing the indicator variable with the largest difference between the
maximum value in a given loading compared to to said indicators loadings
in all other latent variables. Each anchor variable was then assigned to
a named group (Canopy Cover, Height and Biomass, and Structural
Complexity).

Following the EFA, we used structural equation modelling, with latent
variables loadings determined by the EFA. We also filtered the loadings
to be greater than or equal to 0.5, allowing us to examine the number
and composition of latent variables found in each ecosystem. Following
this, we predicted the DHIs in a single SEM with covariances between the
DHIs. If a forest structural attribute did not end up in a latent
variable, it was included in the structural equation model as a DHI
predictor, without a latent variable as a mediator.

@@

## Path Analysis

```{r}
#| label: fig-pathbar-prop
#| fig-cap: Proportion of strongest signficant forest structural predictors of yearly productivity metrics in models based on canopy cover vs canopy height.
#| warning: false
#| echo: false

knitr::include_graphics(here::here("outputs", "path_bars.png"))
```

```{r}
#| label: fig-pathbar-pred
#| fig-cap: Bar plot of strongest predictors in path analysis for each BEC zone and predicted DHI variable.
#| warning: false
#| echo: false

knitr::include_graphics(here::here("outputs", "strongest_predictor_bar.png"))
```

```{r}
#| label: fig-pathmap
#| fig-cap: Map of strongest structural attribute predictor strength on yearly productivity metrics by BEC zone for the two proposed models. If there are no significant predictors, shown in dark grey
#| warning: false
#| echo: false

knitr::include_graphics(here::here("outputs", "path_map_edit.png"))
```

## Exploratory Structural Equation Modelling

I also want to run exploratory SEM with and without the DHI's to see
what grouping it goes in. MinDHI has been breaking because some zones
only have 0 as mindhi (makes sense, they are in the alpine).

```{r}
#| label: fig-esem-latent
#| fig-cap: Groups of latent variables with loadings greater than or equal to 0.5 for each BEC zone. Loadings were determined using exploratory factor analysis with up to four latent variables, and selecting the number of possible latent variables with the lowest AIC. All forest structure variables were potentially included as indicators in each latent variable.
#| warning: false
#| echo: false

knitr::include_graphics(here::here("outputs", "latent_groups.png"))
```

Exploratory factor analysis shows three groups of latent variables
(@fig-esem-latent). We assigned the groups to latent variables
indicating canopy cover / biomass metrics, canopy height / biomass
metrics, and structural complexity metrics. Indicators to latent
variable groups by examining anchors for each latent variable. The
anchors were calculated by choosing the indicator variable with the
largest difference between the maximum value in a given loading compared
to to said indicators loadings in all other latent variables.

Both the Ponderosa Pine (PP) and Sub-boreal Pine -- Spruce zones have
one latent variable. Coastal Western Hemlock (CWH), Engelmann Spurce --
Subalpine Fir (ESSF), and Montane Spruce (MS) are the only zones to have
structural complexity based latent variables. CWH is the only zone to
include elevation standard deviation in the structural complexity latent
variable.

```{r}
#| label: fig-esem-latent-dhi
#| fig-cap: Groups of latent variables with loadings greater than or equal to 0.5 for each BEC zone, including the DHIs. Loadings were determined using exploratory factor analysis with up to five latent variables, and selecting the number of possible latent variables with the lowest AIC. All forest structure variables were potentially included as indicators in each latent variable.
#| warning: false
#| echo: false

knitr::include_graphics(here::here("outputs", "latent_dhis.png"))
```

Exploratory factor analysis shows three groups of latent variables. We
assigned these groups to attributes based on the indicator variables in
each latent variable, indicating canopy cover, height and biomass, and
structural complexity. When including the DHIs in this EFA, it is rare
that they are included with latent variables with large loadings. The
cumulative DHI is an indicator for the canopy cover latent varible in
the IDF (Interior Douglas-fir) and MS (Montane Spruce) BEC zones. The
DHIs are not included in any other latent variable, indicating that they
compose different information than the forest structural variables.

```{r}
#| label: fig-latent-boxplots
#| fig-cap: Boxplots of latent variable loadings across BEC zones as determined by exploratory factor analysis. Red dashed lines show latent variable indicator cutoff before being used as starting values for the SEM models.
#| warning: false
#| echo: false

knitr::include_graphics(here::here("outputs", "latent_boxplot.png"))
```

@fig-latent-boxplots shows boxplots of latent variable loadings from
forest structure indicator variables. A maximum of three valid latent
variables were identified across the sixteen BEC zones. We assigned
indicators variables to latent variable groups by examining the
anchoring variables for each latent variable. The anchoring variables
were calculated by choosing the indicator variable with the largest
difference between the maximum value in a given loading compared to said
indicators loadings in all other latent variables. The only DHI variable
that was included in the latent variables after subsetting the latent
variables to only include strong loadings was the Cumulative DHI in the
canopy cover latent variable.

I need to actually make the DAGs and plot the parameters, but I haven't
gotten there yet. I'll plot the global one to show what it looks like,
then aggregate it in some way - probably similar to the path analysis. I
also may include the variable partitioning that Meg talked about in the
email if I can get it working.
